---
title: "Demonstration of HyperTraPS in R"
output: html_document
---

HyperTraPS (hypercubic transition path sampling) is a family of algorithms for inferring the dynamics of "accumulation" processes. These are processes where a set of binary features are acquired over time.

HyperTraPS will take a set of observed states, described by binary strings recording the presence/absence of each feature. It may optionally take initial states from which these states are reached, and information on the timings associated with each observation. It returns various summaries of which feature are acquired and when, and how these features influence each other.

### Loading the software

If we just want the HyperTraPS code without other dependencies, we only need Rcpp, and can use the following

```{r}
library(Rcpp)
sourceCpp("hypertraps-r.cpp")
```

If we want various helper functions and ggplot functions in addition, use this

```{r}
source("hypertraps.R")
```

### Simple demo 

Here we'll construct a simple synthetic dataset. The `m.1` matrix will store a set of initial states, and the `m.2` matrix will store a set of later observed states. The first row of `m.1`, for example, stores the state 00000, where no features have been acquired. The first row of `m.2` stores 10000, where the first of five features has been acquired.

The times correspond to each observation, so the transition 00000->10000 described by those first rows has an associated time of 0.1 (in whatever units we are working with).

```{r}
m.1 = matrix(rep(c(0,0,0,0,0,
               1,0,0,0,0,
               1,1,0,0,0,
               1,1,1,0,0,
               1,1,1,1,0,
               0,0,0,0,0,
               0,0,0,0,1,
               0,0,0,1,1,
               0,0,1,1,1,
               0,1,1,1,1),5), byrow=TRUE, ncol=5)
m.2 = matrix(rep(c(1,0,0,0,0,
               1,1,0,0,0,
               1,1,1,0,0,
               1,1,1,1,0,
               1,1,1,1,1,
               0,0,0,0,1,
               0,0,0,1,1,
               0,0,1,1,1,
               0,1,1,1,1,
               1,1,1,1,1),5), byrow=TRUE, ncol=5)
times = rep(c(0.1, 0.2, 0.3, 0.4, 0.5), 10)
```

Let's run HyperTraPS with these "before" and "after" observations. By using `times` as both the start and end time arguments, we say that each transition takes precisely that associated time -- we could allow broader time windows to capture uncertainty in timing. Finally, we provide labels for the five individual features involved.

```{r}
my.post = HyperTraPS(m.2, initialstates = m.1, 
                     starttimes = times, endtimes = times,
                     featurenames = c("A", "B", "C", "D", "E")) 
```

That output takes us through the HyperTraPS process. First, the arguments provided to the function call are described, then the algorithm chosen (MH MCMC) and a summary of the input data is given. To estimate runtime, HyperTraPS reports the time taken for a single likelihood calculation, then scales this by the estimated number of calculations required to give a runtime estimate. Here this is only 1.17 seconds, but for chains long enough for satisfactory convergence (and more complicated systems) this may be dramatically longer.

Because we didn't turn off output, HyperTraPS periodically outputs information about the run as it goes. Every 100th step, it outputs: the likelihood of the current parameterisation; the step acceptance rate through the whole run; the step acceptance rate since the last output; the likelihood of the most recent proposal. These can help design efficient MCMC approaches: if the acceptance rate is too low and/or the recent proposal likelihood is much lower than the current, consider a smaller perturbation kernel. It also helps us see when the chain is "burned in" -- when the likelihood fluctuates, rather than consistently increasing, we're probably more stable.

After the MCMC run, the posterior analysis begins, and outputs a few details. These include the size of the posterior sample set being explored, the model being used, and a quick summary of the mean acquisition orderings of each feature. These are really just checks for debugging; not much useful information can be seen from them.

Now -- visualising the results.

*Transition graph plot.* This plot shows a set of sampled routes on the hypercubic space. Each edge is a transition, corresponding to the acquisition of a new feature. The edges are labelled by the feature acquired and the statistics of the time taken (mean +- s.d.). The edge width gives the probability flux through that transition.

```{r}
plotHypercube.sampledgraph2(my.post)
```

*Bubble plot.* This plot summarises the inferred dynamics of features, forgetting specific states. The size of a circle at point x,y gives the probability that feature y is acquired at ordinal time x.

```{r}
plotHypercube.bubbles(my.post)
```

*Timing histograms.* Histograms of acquisition time for each feature, and probability that each feature is not acquired within a given time threshold (here, 20).

```{r}
plotHypercube.timehists(my.post)
```

*Time series plot.* Each step up the vertical axis corresponds to the acquisition of a new feature. The horizontal axis gives the time of the corresponding event, and the colour labels the feature that was acquired.

```{r}
plotHypercube.timeseries(my.post)
```

*Influences between features.* How each acquired trait (horizontal axis) influences the propensity for each other trait (vertical axis) to be acquired. Red is repression, blue is promotion; the transparency gives the width of the associated posterior. Here we see that 1 and 5 strongly cross-repress (being the first steps on the two competing pathways) and each feature promotes the next step on its pathway (wrapping the diagonal).

```{r}
plotHypercube.influences(my.post)
```

### Predictions of unseen and future behaviour

We can use the inferred transition network to make predictions about what will happen next in a given state. Let's query the inferred hypercube to see what will happen next when we are part of the way down one pathway.

```{r}
prediction.step = predictNextStep(my.post, c(1,1,0,0,0))
plotHypercube.prediction(prediction.step)
```

The graphic reports the likely next steps, as a word cloud and a set of probabilities. Here, we're making the strong prediction that the next step after 11000 is 11100.

We can also make predictions about hidden values in new observations. First let's mask an observation and ask HyperTraPS to "fill in the blanks" of 1????.

```{r}
prediction.hidden = predictHiddenVals(my.post, c(1,2,2,2,2))
plotHypercube.prediction(prediction.hidden)
``` 

Here, we see a word cloud of possible states that could correspond to our observation, and a set of probabilities associated with each *feature* being 1. That is, it's highly likely that feature 2 is 1, and less likely that feature 5 is 1. But doesn't this assume something about how far we've come on the hypercube?

Yes -- by default this assumes that all "levels" of the hypercube -- all counts of "1" that are possible, given our uncertainty -- are equally likely. But we might be more interested in specifying something about how many features we're likely to have acquired. We can do this by specifying weights for each "level":

```{r}
prediction.hidden = predictHiddenVals(my.post, c(1,2,2,2,2), level.weight=c(0,0,1,0,0,0))
plotHypercube.prediction(prediction.hidden)
```

Here, we're saying that we believe there's only one more "1" in the ?s -- and we see a correspondingly strong prediction about where that "1" is. We can be more agnostic, and the predictions are shaped accordingly, like below. We might want our prediction to mirror how many "levels" we saw in our original data, for example.

```{r}
predictHiddenVals(my.post, c(1,2,2,2,2), level.weight=c(0,0,1,1,1,0))
plotHypercube.prediction(prediction.hidden)
```

### Uncertainty in observation times

HyperTraPS-CT allows uncertainty in transition timings to be captured. In the run above, we specified a precise timing for each transition, by giving a zero-width time window for each observation (start times = end times). This time window is flexible. We can allow it to have infinite width, in which case absolute timings are meaningless and we just infer the orderings of events. Or we can specify a finite time window, allowing a transition to take any time within that window, to capture uncertainty in observation timings.

```{r}
# precisely specified timings, as above
my.post.time.precise = HyperTraPS(m.2, initialstates = m.1, 
                               starttimes = times, endtimes = times,  
                               limited_output = 1,
                               featurenames = c("A", "B", "C", "D", "E")); 
# infinite width time window for transitions (equivalent to just inferring ordering)
my.post.time.inf = HyperTraPS(m.2, initialstates = m.1, 
                                starttimes = times*0, endtimes = times*Inf, 
                                limited_output = 1,
                                featurenames = c("A", "B", "C", "D", "E"));
# finite time window for each uncertain transition time
my.post.time.uncertain = HyperTraPS(m.2, initialstates = m.1, 
                     starttimes = times*0.25, endtimes = times*4, 
                     limited_output = 1,
                     featurenames = c("A", "B", "C", "D", "E")); 
ggarrange(plotHypercube.timehists(my.post.time.precise, t.thresh=3), 
          plotHypercube.timehists(my.post.time.uncertain, t.thresh=3),
          plotHypercube.timehists(my.post.time.inf, t.thresh=3),
          nrow=3)
plotHypercube.sampledgraph2(my.post.time.precise, thresh=0.1, use.arc=FALSE, edge.label.size=3) + theme(legend.position="none") + expand_limits(x = c(-0.1, 1.1))
plotHypercube.sampledgraph2(my.post.time.uncertain, thresh=0.1, use.arc=FALSE, edge.label.size=3) + theme(legend.position="none") + expand_limits(x = c(-0.1, 1.1))
plotHypercube.sampledgraph2(my.post.time.inf, thresh=0.1, use.arc=FALSE, edge.label.size=3) + theme(legend.position="none") + expand_limits(x = c(-0.1, 1.1))
```

The summary plots here reflects the different inferences about acquisition timescales that come from the approaches with different time windows. 

### Input/output between R data structure and file format

We can write the output of HyperTraPS to a set of files for storage and later retrieval

```{r}
writeHyperinf(my.post, "simpledemo", my.post$L, postlabel = "simpledemo", fulloutput=TRUE)
```

We can retrieve output from files, which may have been previously written as above, or may have come from running HyperTraPS at the command line.

```{r}
my.post.r = readHyperinf("simpledemo", postlabel = "simpledemo", fulloutput=TRUE)
```

### Other functional examples

HyperTraPS allows substantial flexibility in how evolutionary pathways are inferred, how likelihoods are estimated, and other aspects of the approach. 

If we want to sacrifice some accuracy in estimating likelihood for computational speed, we can run an example with fewer walkers sampling pathways (`walkers`)

```{r}
my.post.sparse = HyperTraPS(m.2, initialstates = m.1, 
                            starttimes = times, featurenames = c("A", "B", "C", "D", "E"), 
                            walkers = 2,
                            limited_output = 1)
```

We can ask HyperTraPS to perform stepwise regularisation after fitting a model, pruning extraneous parameters (`regularise`). The regularisation plot shows how the information criterion behaves as we prune back parameters -- the minimum gives us our optimal model.

```{r}
my.post.regularise = HyperTraPS(m.2, initialstates = m.1, regularise = 1,
                                walkers = 20,
                                limited_output = 1)
plotHypercube.regularisation(my.post.regularise)
plotHypercube.summary(my.post.regularise)
```

We can use simulated annealing to get a maximum likelihood estimate rather than a Bayesian picture (`sa`)

```{r}
my.post.sa = HyperTraPS(m.2, initialstates = m.1, sa = 1,
                        limited_output = 1)
plotHypercube.summary(my.post.sa)
```

We can also use "phenotype landscape inference" -- an unbiased sampling method, unlike the (corrected) bias sampling in HyperTraPS (`PLI`). This takes longer but is more flexible with uncertain data

```{r}
my.post.pli = HyperTraPS(m.2, initialstates = m.1, pli = 1,
                         limited_output = 1)
plotHypercube.summary(my.post.pli)
```

HyperTraPS supports different parameter structures (`model`). By default we use an "L2" parameterisation, where each feature can individually influence the acquisition of every other feature. "L1" has features completely independent; "L0" has all features identical. "L3" allows *pairs* of features to influence each feature's acquisition; "L4" allows *triplets* of features to influence each feature's acquisition. This approach, labelled -1, allows every edge on the hypercube to have its own independent parameters -- corresponding to the most flexible possible picture, where arbitrary sets of features influence other features. We then regularise as above (`regularise`) to remove extraneous parameters

```{r}
my.post.bigmodel.regularise = HyperTraPS(m.2, initialstates = m.1, model = -1,
                                         regularise = 1, walkers = 20,
                                         limited_output = 1)
plotHypercube.regularisation(my.post.bigmodel.regularise)
```

Here's another example of model choice. The data is now generated by a process where *pairs* of features influence the acquisition of other features. To capture these interactions, an "L^2" picture (`model = 2`, where each feature *individually* influences each other feature) is insufficient -- an "L^3" picture (`model = 3`) allowing pair influence is needed. The full parameterisation, where every edge on the hypercube transition network (`model = -1`) has an independent parameter, can also capture this behaviour.

```{r}
logic.mat = readLines("Verify/hi-order.txt")
logic.mat = do.call(rbind, lapply(strsplit(logic.mat, ""), as.numeric))
logic.starts = logic.mat[seq(from=1, to=nrow(logic.mat), by=2),]
logic.ends = logic.mat[seq(from=2, to=nrow(logic.mat), by=2),]
logic.post.m1 = HyperTraPS(logic.ends, initialstates = logic.starts, length = 4, model = -1, walkers = 20, limited_output = 1)
logic.post.1 = HyperTraPS(logic.ends, initialstates = logic.starts, length = 4, model = 1, walkers = 20, limited_output = 1)
logic.post.2 = HyperTraPS(logic.ends, initialstates = logic.starts, length = 4, model = 2, walkers = 20, limited_output = 1)
logic.post.3 = HyperTraPS(logic.ends, initialstates = logic.starts, length = 4, model = 3, walkers = 20, limited_output = 1)

ggarrange(plotHypercube.graph(logic.post.m1) + ggtitle("All edges") + theme(legend.position="none"),
          plotHypercube.graph(logic.post.1) + ggtitle("L") + theme(legend.position="none"),
          plotHypercube.graph(logic.post.2)+ ggtitle("L^2") + theme(legend.position="none"),
          plotHypercube.graph(logic.post.3)+ ggtitle("L^3") + theme(legend.position="none"))
```    

We can see that the structures inferred by the "full" and "L^3" models are the same, while the "L^2" (and inappropriate "L^1") will either introduce extraneous transitions or fail to capture the true ones.

We can also impose priors on our model. This is achieved by specifying ranges for each parameter. Here, we give an extreme example, constructing a matrix setting this range to only include zero for all parameters except basal acquisition rates.

```{r}
priors = matrix(0, ncol=2, nrow=5*5)
for(i in 0:4) {
  priors[i*5+i+1,1] = -10
  priors[i*5+i+1,2] = 10
}
my.post.priors = HyperTraPS(m.2, initialstates = m.1, 
                     starttimes = times, endtimes = times, 
                     priors = priors,
                     limited_output = 1,
                     featurenames = c("A", "B", "C", "D", "E")); 
plotHypercube.summary(my.post.priors)
```

As this prior structure disallows interactions between features (these are exactly the off-diagonal parameters that we've enforced to be zero), the results should resemble inference run using the reduced model structure without interactions:

```{r}
my.post.model1 = HyperTraPS(m.2, initialstates = m.1, 
                            starttimes = times, endtimes = times, 
                            model = 1, kernel=3,
                            limited_output = 1,
                            featurenames = c("A", "B", "C", "D", "E")); 
plotHypercube.summary(my.post.model1)
```

### Scientific examples

A set of short-form examples from past studies -- these should run in a few minutes and give approximations to the original results. In each case we read the observed states from a file (these are present in different formats, so different curation steps are involved in each case), and feature labels from another file. Then we run HyperTraPS and plot some summaries of the output.

1. Ovarian cancer case study: traits are chromosomal aberrations, observations are independent patient samples.

```{r}
cgh.mat = readLines("RawData/ovarian.txt")
cgh.mat = do.call(rbind, lapply(strsplit(cgh.mat, ""), as.numeric))
cgh.names = as.vector(read.table("RawData/ovarian-names.txt", sep=","))[[1]]

my.post.cgh = HyperTraPS(cgh.mat, 
                        length = 3, outputinput = 1, 
                        featurenames = cgh.names,
                        limited_output = 1) 
ggarrange(plotHypercube.lik.trace(my.post.cgh), 
          plotHypercube.bubbles(my.post.cgh, reorder=TRUE), 
          plotHypercube.sampledgraph2(my.post.cgh, no.times=TRUE), nrow=3)
plotHypercube.sampledgraph2(my.post.cgh, no.times=TRUE)
```

2. C4 photosynthesis case study: traits are physical/genetic features associated with C4, observations are (incomplete) phylogenetically independent intermediate species.

This case study involves some uncertain data, which are present as "2"s in the source data -- labelling features which may be 0 or 1.

```{r}
c4.mat = as.matrix(read.table("RawData/c4-curated.csv", sep=","))
c4.names = as.vector(read.table("RawData/c4-trait-names.txt", sep=","))[[1]]

my.post.c4 = HyperTraPS(c4.mat, 
                        length = 3, 
                        losses = 1,
                        featurenames = c4.names,
                        limited_output = 1) 
plotHypercube.bubbles(my.post.c4, reorder=TRUE)
```

3. Severe malaria disease progression case study: traits are clinical features, observations are (incomplete) independent patient presentations

```{r}
malaria.df = read.csv("RawData/jallow_dataset_binary_with2s.csv")
malaria.mat = as.matrix(malaria.df[,2:ncol(malaria.df)])
malaria.names = as.vector(read.table("RawData/malaria-names.txt", sep=","))[[1]]

my.post.malaria = HyperTraPS(malaria.mat, 
                        length = 3,
                        kernel = 2,
                        walkers = 20,
                        featurenames = malaria.names,
                        limited_output = 1) 
plotHypercube.bubbles(my.post.malaria, reorder=TRUE, transpose=TRUE)
```

4. Tool use evolution case study: traits are modes of tool use, observations are phylogenetically coupled species observations (phylogeny has been accounted for, giving transition pairs)

```{r}
tools.mat = as.matrix(read.table("RawData/total-observations.txt-trans.txt"))
tools.names = as.vector(read.table("RawData/tools-names.txt"))[[1]]
tools.starts = tools.mat[seq(from=1, to=nrow(tools.mat), by=2),]
tools.ends = tools.mat[seq(from=2, to=nrow(tools.mat), by=2),]

my.post.tools = HyperTraPS(tools.ends, initialstates = tools.starts, 
                           length = 3, 
                           featurenames = tools.names,
                           limited_output = 1) 
plotHypercube.bubbles(my.post.tools, reorder=TRUE, transpose=TRUE)
plotHypercube.sampledgraph2(my.post.tools, node.labels = FALSE, max=100, no.times=TRUE) + theme(legend.position = "none")
```

